---
title: "QUILT2 parameter validation"
subtitle: "Overall reported GT accuracy and its associations with INFO_SCORE and HWE"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
params:
  data_file: null
---

```{r}
#| label: setup
suppressPackageStartupMessages({
  library(arrow)
  library(dplyr)
  library(ggplot2)
  library(jsonlite)
  library(knitr)
  library(scales)
})

if (is.null(params$data_file) || !nzchar(params$data_file)) {
  stop("Set params$data_file to quilt2_parameter_extract.parquet.", call. = FALSE)
}
data_file <- normalizePath(params$data_file, mustWork = TRUE)
if (!grepl("[.]parquet$", data_file, ignore.case = TRUE)) {
  stop("params$data_file must be a Parquet file.", call. = FALSE)
}

reader <- ParquetFileReader$create(data_file)
schema <- reader$GetSchema()
parquet_metadata <- schema$metadata
dataset <- open_dataset(data_file, format = "parquet")

required_columns <- c(
  "locus_id", "chrom", "pos", "run_id", "treatment", "panel", "truth_source",
  "sample_id", "array_group", "mean_info_score", "info_plot_eligible",
  "percent_correct_pooled", "pooled_site_eligible", "n_expected_pooled",
  "n_valid_pooled", "n_correct_pooled", "info__hwe", "hwe_valid",
  "percent_correct_panel", "site_cc_eligible", "n_expected_panel",
  "n_valid_panel", "n_correct_panel"
)
missing_columns <- setdiff(required_columns, names(schema))
if (length(missing_columns)) {
  stop("Parquet is missing required columns: ", paste(missing_columns, collapse = ", "), call. = FALSE)
}
if (!identical(parquet_metadata$extraction_schema_version, "quilt2-parameter-validation-v4")) {
  stop("This report requires v4 output with the twelve-run Array/WGS evaluation mask.", call. = FALSE)
}
required_metadata <- c(
  "accuracy_target", "analysis_compromise",
  "eval_mask_position_count", "eval_mask_sha256",
  "eval_mask_manifest_sha256", "eval_mask_manifest_json",
  "common_locus_count", "final_mask_attrition_from_eval", "mask_sha256",
  "wgs_truth_min_gq", "wgs_truth_min_dp", "gp_sum_tolerance", "source_manifest_json"
)
missing_metadata <- required_metadata[!vapply(required_metadata, function(key) {
  !is.null(parquet_metadata[[key]])
}, logical(1))]
if (length(missing_metadata)) {
  stop("Parquet is missing required metadata: ", paste(missing_metadata, collapse = ", "), call. = FALSE)
}
if (!identical(parquet_metadata$accuracy_target, "reported_gt")) {
  stop("Accuracy must refer to the reported GT call.", call. = FALSE)
}

treatment_levels <- c("Filtered", "No_filter")
treatment_labels <- c(Filtered = "Filtered", No_filter = "No-filter")
theme_set(theme_bw(base_size = 11))
theme_report <- theme(
  panel.grid.minor = element_blank(),
  strip.background = element_rect(fill = "grey94", colour = "grey75"),
  legend.position = "right"
)

assert_unique <- function(data, keys, label) {
  duplicates <- data |>
    count(across(all_of(keys)), name = "n") |>
    filter(n != 1L)
  if (nrow(duplicates)) stop(label, " is not unique at its declared grain.", call. = FALSE)
}

assert_true <- function(value, message) {
  if (!isTRUE(value)) stop(message, call. = FALSE)
}

format_metadata <- function(key, fallback = NA_character_) {
  value <- parquet_metadata[[key]]
  if (is.null(value)) fallback else as.character(value)
}
```

## Scope and interpretation

Agreed compromise: this report asks how often the reported GT calls are correct,
regardless of their GP confidence. It does not assess probability calibration,
GP-argmax correctness, or whether a confidence threshold improves accuracy.
GP bins, reliability curves, calibration-gap heatmaps, and GT-versus-GP-argmax
comparisons have been removed. Missing, malformed, or tied GP does not exclude
an otherwise evaluable GT call. Valid aligned truth is still required.

This report evaluates whether QUILT2's emitted parameters describe observed
imputation behaviour at the common loci. It does not estimate a causal filter
effect. The starting position mask is the CHROM/POS intersection of loci present
in all six completed Array concordance Parquets and all six WGS
`per_variant_metrics` Parquet datasets. Their paths are fixed in the extraction
script. Missing concordance values or incomplete valid-call counts do not remove
a position from the mask; presence is separate from call-level evaluability.
The same position mask is applied to the Array and WGS runs and both truth
resources. The extractor then requires compatible records in all physical
QUILT2 outputs and both truth resources, so the plotted exact-allele mask can be
a subset of the twelve-run evaluator position mask.

The 25 targets comprise 18 Array Group I/II samples from the 2× run and seven
WGS samples. Reference-panel identity is retained in the Parquet for auditing,
but it is not represented by plot colour, shape, or facets.

Array and WGS targets were imputed in separate physical runs. Consequently,
each treatment–panel condition has two emitted INFO_SCORE and HWE values. The
18 Array FORMAT columns are selected for call-level evaluation, but site-level
fields remain exactly as emitted by the original physical run; they are not
recomputed after sample selection. The embedded source manifest records the
original sample count and a hash of its complete sample list so this scope is
auditable. The
extractor defines the plotted `mean_info_score` as the target-count-weighted
mean of the 18-sample Array and 7-sample WGS emissions within each panel,
followed by an equal mean over the three panels. HWE p-values are never averaged:
both physical emissions enter the HWE analysis as separate observations paired
with the same 25-sample panel correctness.

```{r}
#| label: provenance
provenance <- tibble::tibble(
  item = c(
    "Parquet", "Schema version", "Rows",
    "Twelve-run Array/WGS evaluator-mask positions", "Evaluator-mask SHA-256",
    "Evaluation-mask manifest SHA-256",
    "Final exact-allele common loci", "Positions removed after evaluator mask",
    "Final mask SHA-256",
    "WGS truth minimum GQ", "WGS truth minimum DP", "GP sum tolerance (audit only)",
    "Accuracy target", "Documented analysis compromise"
  ),
  value = c(
    data_file,
    format_metadata("extraction_schema_version"),
    comma(reader$num_rows),
    comma(as.numeric(format_metadata("eval_mask_position_count"))),
    format_metadata("eval_mask_sha256"),
    format_metadata("eval_mask_manifest_sha256"),
    comma(as.numeric(format_metadata("common_locus_count"))),
    comma(as.numeric(format_metadata("final_mask_attrition_from_eval"))),
    format_metadata("mask_sha256"),
    format_metadata("wgs_truth_min_gq"),
    format_metadata("wgs_truth_min_dp"),
    format_metadata("gp_sum_tolerance"),
    format_metadata("accuracy_target"),
    format_metadata("analysis_compromise")
  )
)
kable(provenance, caption = "Embedded extraction provenance")

eval_mask_sources <- jsonlite::fromJSON(format_metadata("eval_mask_manifest_json")) |>
  as_tibble() |>
  transmute(
    run_id = logical_run_id,
    treatment,
    panel,
    truth_source,
    source_type = mask_input_type,
    retained_positions = position_count,
    artifact_sha256_prefix = substr(source_sha256, 1, 12),
    position_sha256_prefix = substr(positions_sha256, 1, 12)
  ) |>
  arrange(factor(treatment, levels = treatment_levels), panel, truth_source)
kable(
  eval_mask_sources,
  caption = "Six Array and six WGS evaluation artifacts used to construct the starting position mask"
)

source_population <- jsonlite::fromJSON(format_metadata("source_manifest_json")) |>
  as_tibble() |>
  transmute(
    run_id = logical_run_id,
    treatment,
    panel,
    target_run = truth_source,
    original_samples = source_sample_count,
    selected_samples = selected_sample_count,
    sample_list_sha256_prefix = substr(source_samples_sha256, 1, 12)
  ) |>
  arrange(factor(treatment, levels = treatment_levels), panel, target_run)
kable(source_population, caption = "Physical-run sample populations (12-character sample-list hash prefix shown)")
```

## Validated analysis views

Only columns needed by a particular view are selected through Arrow. The full
call table is not loaded as one local object. Overall GT accuracy reuses the
precomputed pooled counts from the unfiltered INFO view, including loci with
missing INFO_SCORE or incomplete calls. GP columns are not read by Quarto.

```{r}
#| label: views
info_view <- dataset |>
  select(
    locus_id, chrom, treatment, mean_info_score, info_plot_eligible,
    percent_correct_pooled, pooled_site_eligible, n_expected_pooled,
    n_valid_pooled, n_correct_pooled
  ) |>
  distinct() |>
  collect() |>
  mutate(treatment = factor(treatment, levels = treatment_levels))

hwe_view <- dataset |>
  select(
    locus_id, chrom, treatment, panel, truth_source, info__hwe, hwe_valid,
    percent_correct_panel, site_cc_eligible, n_expected_panel,
    n_valid_panel, n_correct_panel
  ) |>
  distinct() |>
  collect() |>
  mutate(treatment = factor(treatment, levels = treatment_levels))

sample_lookup <- dataset |>
  select(sample_id, array_group) |>
  distinct() |>
  collect() |>
  arrange(factor(array_group, levels = c("Group I", "Group II", "WGS")), sample_id)

assert_unique(info_view, c("locus_id", "treatment"), "INFO view")
assert_unique(hwe_view, c("locus_id", "treatment", "panel", "truth_source"), "HWE view")
assert_unique(sample_lookup, "sample_id", "Sample lookup")
assert_true(nrow(sample_lookup) == 25L, "Expected exactly 25 samples in the Parquet.")
assert_true(
  identical(sort(unique(as.character(info_view$treatment))), sort(treatment_levels)),
  "Expected both Filtered and No_filter treatments."
)
assert_true(all(info_view$n_expected_pooled == 75L), "n_expected_pooled must always be 75.")
assert_true(all(hwe_view$n_expected_panel == 25L), "n_expected_panel must always be 25.")
assert_true(
  all(info_view$n_correct_pooled >= 0 &
    info_view$n_correct_pooled <= info_view$n_valid_pooled &
    info_view$n_valid_pooled <= info_view$n_expected_pooled),
  "Pooled counts are inconsistent, including at incomplete loci."
)
assert_true(
  all(is.na(info_view$percent_correct_pooled[!info_view$pooled_site_eligible])),
  "Ineligible pooled sites must have percent_correct_pooled = NA."
)
assert_true(
  all(is.na(hwe_view$percent_correct_panel[!hwe_view$site_cc_eligible])),
  "Ineligible panel sites must have percent_correct_panel = NA."
)
pooled_check <- info_view |>
  filter(pooled_site_eligible) |>
  transmute(ok = n_valid_pooled == 75L &
    abs(percent_correct_pooled - 100 * n_correct_pooled / 75) < 1e-10)
panel_check <- hwe_view |>
  filter(site_cc_eligible) |>
  transmute(ok = n_valid_panel == 25L &
    abs(percent_correct_panel - 100 * n_correct_panel / 25) < 1e-10)
assert_true(all(pooled_check$ok), "Supplied pooled %CC fails its fixed-denominator invariant.")
assert_true(all(panel_check$ok), "Supplied panel %CC fails its fixed-denominator invariant.")
view_summary <- tibble::tibble(
  view = c("INFO and overall GT counts", "HWE"),
  grain = c(
    "locus × treatment",
    "locus × treatment × panel × physical target run"
  ),
  rows = c(nrow(info_view), nrow(hwe_view))
)
kable(view_summary, digits = 0, caption = "In-memory analysis views")
```

## `%CC` versus `INFO_SCORE`

Why use this graph: `INFO_SCORE` is a site-level imputation-quality statistic,
so the relevant validation question is whether it ranks loci from lower to
higher observed hard-call accuracy. It is not a correctness probability; a
1:1 line would therefore imply the wrong calibration interpretation.

What it shows: each point is one eligible locus × treatment. Its y-value is the
precomputed percentage of correct GT calls across 75 comparisons (25 targets ×
three panels). The 0.05-wide summaries expose the broad trend while preserving
the locus cloud. Confidence bands are chromosome-block jackknife intervals and
therefore respect within-chromosome dependence better than ordinary locus-wise
standard errors.

```{r}
#| label: info-prepare
info_breaks <- c(seq(0, 0.95, by = 0.05), 1 + .Machine$double.eps)
info_labels <- sprintf("%.2f–%.2f", head(info_breaks, -1), pmin(tail(info_breaks, -1), 1))

info_plot_data <- info_view |>
  filter(info_plot_eligible, is.finite(mean_info_score), !is.na(percent_correct_pooled)) |>
  mutate(
    info_bin = cut(
      mean_info_score,
      breaks = info_breaks,
      labels = info_labels,
      include.lowest = TRUE,
      right = FALSE
    )
  ) |>
  filter(!is.na(info_bin))

jackknife_bin <- function(data) {
  blocks <- unique(data$chrom)
  full_estimate <- mean(data$percent_correct_pooled)
  leave_one_out <- vapply(
    blocks,
    function(block) mean(data$percent_correct_pooled[data$chrom != block]),
    numeric(1)
  )
  leave_one_out <- leave_one_out[is.finite(leave_one_out)]
  block_count <- length(leave_one_out)
  if (block_count >= 2L) {
    leave_one_out_mean <- mean(leave_one_out)
    standard_error <- sqrt((block_count - 1) / block_count *
      sum((leave_one_out - leave_one_out_mean)^2))
    lower <- max(0, full_estimate - 1.96 * standard_error)
    upper <- min(100, full_estimate + 1.96 * standard_error)
  } else {
    standard_error <- lower <- upper <- NA_real_
  }
  tibble::tibble(
    mean_info_score = mean(data$mean_info_score),
    percent_correct = full_estimate,
    n_loci = nrow(data),
    n_chromosomes = length(blocks),
    jackknife_se = standard_error,
    lower = lower,
    upper = upper
  )
}

info_summary <- info_plot_data |>
  group_by(treatment, info_bin) |>
  group_modify(~ jackknife_bin(.x)) |>
  ungroup()
```

```{r}
#| label: fig-info-score
#| fig-cap: "INFO_SCORE rank validation. Points are eligible loci; lines are 0.05-bin means and ribbons are chromosome-block jackknife 95% intervals."
#| fig-width: 10
#| fig-height: 5
ggplot(info_plot_data, aes(mean_info_score, percent_correct_pooled)) +
  geom_point(alpha = 0.12, size = 0.7, colour = "#24527A") +
  geom_ribbon(
    data = info_summary,
    aes(x = mean_info_score, ymin = lower, ymax = upper),
    inherit.aes = FALSE,
    fill = "#E69F00",
    alpha = 0.22
  ) +
  geom_line(
    data = info_summary,
    aes(x = mean_info_score, y = percent_correct),
    inherit.aes = FALSE,
    colour = "#B24C00",
    linewidth = 0.9
  ) +
  geom_point(
    data = info_summary,
    aes(x = mean_info_score, y = percent_correct),
    inherit.aes = FALSE,
    colour = "#B24C00",
    size = 1.8
  ) +
  facet_wrap(~ treatment, labeller = as_labeller(treatment_labels)) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_y_continuous(limits = c(0, 100), labels = label_percent(scale = 1)) +
  labs(
    x = "Mean QUILT2 INFO_SCORE",
    y = "Correct GT calls across 75 comparisons (%CC)",
    caption = "Descriptive common-locus association; no causal treatment claim and no 1:1 line."
  ) +
  theme_report
```

```{r}
#| label: info-summary-table
kable(
  info_summary |>
    mutate(treatment = recode(as.character(treatment), !!!treatment_labels)) |>
    select(treatment, info_bin, mean_info_score, percent_correct, n_loci, n_chromosomes, lower, upper),
  digits = 3,
  caption = "INFO_SCORE-bin summaries and chromosome-block jackknife intervals"
)
```

## Overall reported GT accuracy

Why use this graph: it answers the practical question, "How accurate are the
reported calls regardless of confidence?" Each evaluable sample × locus × panel
GT contributes one correct or incorrect comparison within its treatment.

What it shows: the percentage of evaluable GT calls matching aligned truth,
pooled across all common-mask loci, 25 target samples, and three panels. It
includes valid comparisons from incomplete loci and from loci with missing
INFO_SCORE. The denominator is the total number of valid GT–truth comparisons,
and the table also reports unevaluable comparisons. This aggregate differs from
site-level `%CC`, which still requires all 25 or 75 calls at an individual site.

Compromise: no GP threshold, confidence bin, or GP-argmax selection is applied.
The plot cannot establish whether GP probabilities are calibrated, whether
high-confidence calls are more accurate, or which GP cutoff should be used.
The panels contribute repeated evaluations of the same targets, so this remains
a descriptive summary without an independence-based significance test.

```{r}
#| label: overall-gt-summary
# Use all precomputed pooled counts, without filtering on site eligibility,
# INFO_SCORE, HWE, or any GP value. No genotype comparisons are performed here.
gt_accuracy_summary <- info_view |>
  group_by(treatment) |>
  summarise(
    n_loci = n_distinct(locus_id),
    n_expected_calls = sum(n_expected_pooled),
    n_valid_calls = sum(n_valid_pooled),
    n_correct_calls = sum(n_correct_pooled),
    .groups = "drop"
  ) |>
  mutate(
    n_incorrect_calls = n_valid_calls - n_correct_calls,
    n_unevaluable_calls = n_expected_calls - n_valid_calls,
    percent_correct = if_else(
      n_valid_calls > 0,
      100 * n_correct_calls / n_valid_calls,
      NA_real_
    )
  )
```

```{r}
#| label: fig-overall-gt-accuracy
#| fig-cap: "Overall reported GT correctness by treatment, regardless of GP confidence. Each point uses all valid GT–truth comparisons in the common mask."
#| fig-width: 7
#| fig-height: 5
gt_accuracy_plot_data <- gt_accuracy_summary |>
  filter(n_valid_calls > 0)

ggplot(gt_accuracy_plot_data, aes(treatment, percent_correct)) +
  geom_point(colour = "#24527A", size = 3.5) +
  geom_text(aes(label = sprintf("%.2f%%", percent_correct)), vjust = -0.8) +
  scale_x_discrete(limits = treatment_levels, labels = treatment_labels, drop = FALSE) +
  scale_y_continuous(
    limits = c(0, 100), labels = label_percent(scale = 1),
    expand = expansion(mult = c(0.02, 0.08))
  ) +
  labs(
    x = "Treatment",
    y = "Correct reported GT calls (%)",
    caption = "Actual valid-call denominator; no GP filtering or calibration assessment."
  ) +
  theme_report
```

```{r}
#| label: overall-gt-table
kable(
  gt_accuracy_summary |>
    mutate(treatment = recode(as.character(treatment), !!!treatment_labels)) |>
    select(
      treatment, n_loci, n_expected_calls, n_valid_calls,
      n_correct_calls, n_incorrect_calls, n_unevaluable_calls, percent_correct
    ),
  digits = 2,
  caption = "Overall GT accuracy and comparison counts, with GP ignored"
)
```

## `%CC` versus HWE

Why use this graph: the emitted HWE p-value describes departure from
Hardy–Weinberg equilibrium at a locus. Relating `-log10(HWE)` to observed
hard-call correctness tests whether stronger departures are associated with
lower accuracy.

What it shows: every y-value is the precomputed 25-sample panel `%CC`. HWE
p-values are physical-run emissions and are not averaged across either panels
or Array/WGS target runs. Panel and target-run identities remain in the data but
are not shown aesthetically. Therefore dense regions contain repeated evidence
from multiple emissions rather than independent biological loci, and the plot
is descriptive rather than an ordinary locus-level significance test.

```{r}
#| label: hwe-prepare
hwe_plot_data <- hwe_view |>
  filter(site_cc_eligible, hwe_valid, !is.na(percent_correct_panel))
positive_hwe <- hwe_plot_data$info__hwe[
  is.finite(hwe_plot_data$info__hwe) & hwe_plot_data$info__hwe > 0
]
if (!length(positive_hwe)) stop("No positive HWE p-values are available for plotting.", call. = FALSE)
hwe_floor <- max(.Machine$double.xmin, min(positive_hwe) / 10)
hwe_plot_data <- hwe_plot_data |>
  mutate(
    hwe_censored = info__hwe == 0,
    hwe_for_plot = if_else(hwe_censored, hwe_floor, info__hwe),
    hwe_departure = -log10(hwe_for_plot),
    hwe_bin_lower = floor(hwe_departure / 0.5) * 0.5
  )

hwe_summary <- hwe_plot_data |>
  group_by(treatment, hwe_bin_lower) |>
  summarise(
    mean_hwe_departure = mean(hwe_departure),
    mean_percent_correct = mean(percent_correct_panel),
    n_emissions = n(),
    n_loci = n_distinct(locus_id),
    .groups = "drop"
  )
```

```{r}
#| label: fig-hwe
#| fig-cap: "HWE departure versus fixed-denominator panel correctness. Lines summarise 0.5-wide -log10(p) bins."
#| fig-width: 10
#| fig-height: 5
ggplot(hwe_plot_data, aes(hwe_departure, percent_correct_panel)) +
  geom_point(aes(shape = hwe_censored), alpha = 0.10, size = 0.75, colour = "#24527A") +
  geom_line(
    data = hwe_summary,
    aes(mean_hwe_departure, mean_percent_correct),
    inherit.aes = FALSE,
    colour = "#B24C00",
    linewidth = 0.9
  ) +
  geom_point(
    data = hwe_summary,
    aes(mean_hwe_departure, mean_percent_correct),
    inherit.aes = FALSE,
    colour = "#B24C00",
    size = 1.7
  ) +
  facet_wrap(~ treatment, labeller = as_labeller(treatment_labels)) +
  scale_shape_manual(
    values = c(`FALSE` = 16, `TRUE` = 17),
    labels = c(`FALSE` = "Positive p-value", `TRUE` = "Reported zero (censored)"),
    name = "HWE value"
  ) +
  scale_y_continuous(limits = c(0, 100), labels = label_percent(scale = 1)) +
  labs(
    x = expression(-log[10]("QUILT2 HWE p-value")),
    y = "Correct GT calls among 25 targets (%CC)",
    caption = paste0(
      "Reported zeros are right-censored and plotted at p = ",
      format(hwe_floor, scientific = TRUE, digits = 3),
      ", one decade below the smallest positive emitted value."
    )
  ) +
  theme_report
```

```{r}
#| label: hwe-summary-table
kable(
  hwe_summary |>
    mutate(treatment = recode(as.character(treatment), !!!treatment_labels)) |>
    select(treatment, hwe_bin_lower, mean_hwe_departure, mean_percent_correct, n_emissions, n_loci),
  digits = 3,
  caption = "HWE-departure-bin summaries"
)
```

## Reading limits

- `%CC` is a hard-call metric. It answers whether the reported GT genotype is
  exactly correct, not how close an allele dosage is.
- Panel `%CC` is reported only for 25/25 valid target calls; pooled `%CC` only
  for 75/75 valid panel–target comparisons. Missing calls are retained in the
  Parquet but do not silently reduce these denominators.
- Overall GT accuracy uses all evaluable GT–truth comparisons, including those
  at incomplete loci; its denominator is their actual count. Unevaluable calls
  are counted separately. GP does not affect scoring or inclusion.
- Agreed compromise: GP confidence is ignored. No confidence bins, GP-argmax
  comparisons, reliability curves, or calibration-gap heatmaps are produced.
  The report cannot assess probability calibration or recommend a GP cutoff.
- The common-locus restriction can narrow the range of each parameter and can
  induce selection effects. The figures validate parameter behaviour within
  this retrospective overlap; they do not isolate a causal filter parameter.
- The starting mask is the intersection of positions present in all six Array
  and six WGS evaluation Parquets. The final plotted mask can be smaller because
  WGS truth and all 12 physical QUILT2 outputs must also pass exact allele and
  uniqueness checks.
